### 2023-03-02 intro.tex

Prior work has separately considered mechanisms for maintaining syntactically well-formed program sketches during development: manual hole insertion (as is featured in Agda \cite{agda}, GHC Haskell \cite{haskell-holes}, Idris \cite{idris-holes}, and other languages), syntax error recovery \cite{error-reovery}, and structure editing \cite{HazelnutPOPL}. 
Hazel has experimented with several of these approaches, offering (1) a textual syntax, which allows the programmer to manually insert empty holes as necessary, (2) a text-like structure editor that offers partial syntax error recovery (inserting holes to maintain well-formedness as long as all matching delimiters have been placed) \cite{tylr}, and (3) a term-based structure editor with total syntax error recovery, i.e. a guarantee that every edit maintains syntactic well-formedness \cite{HazelnutPOPL}\todo{popl17}. The choice of editing method is independent of the mechanisms described in Sec.~\ref{sec:calculus}. However, as a secondary contribution, 
we demonstrate in Sec.~\ref{sec:calculus-structured-editing} that total type error recovery can resolve the problem of undefined editing behavior in the Hazelnut term-based structure editor calculus, which arise because the system as formulated cannot handle edits that would require marking anywhere other than at the cursor. We discuss two reformulations of Hazelnut, first as an untyped structure editor where full re-marking occurs after each edit, then as a bidirectionally typed structure editor where re-marking occurs only where necessary based on the edit location and the type and binding structure.

Local type inference pairs well with type error localization and recovery because information flows systematically down the tree, leading to predictable and systematic localization decisions (as informally observed also by \cite{pierce}). Languages with constraint-based type inference gather constraints from all relevant locations in a program. This approach is powerful, but it is also notorious for complicating type error localization and recovery, because inconsistencies can arise through a confluence of constraints originating from any number of locations in the program. 
Complex techniques, e.g. based on machine learning or manual weighting, have been proposed to heuristically localize errors to expressions. 
In Sec.~\ref{sec:thi}\todo{sec}, we describe how we blend local and constraint-based type inference via type holes in Hazel. In particular, 
we generate constraints for each type hole that appears in the program explicitly or that is  generated internally during the bidirectional type error localization and recovery process. 
We develop a unification algorithm that can recover from inconsistent constraints by maintaining possible type hole fillings. For holes that can be uniquely solved, we suggest the solution. 
If conflicting constraints arise, we localize the error directly to the relevant type holes, rather 
than attempting to localize it to expressions. 
The user can select from different partial fillings, which fills the type hole. From there, the bidirectional system shifts the error to relevant expressions.
Until a choice is made, the system continues to recover gradually, as covered in Sec.~\ref{sec:calculus}.

>>>>

Prior work has separately considered mechanisms for maintaining syntactically well-formed program sketches during development: manual hole insertion (as is featured in Agda \cite{agda}, GHC Haskell \cite{haskell-holes}, Idris \cite{idris-holes}, and other languages), syntax error recovery \cite{error-reovery}, and structure editing \cite{HazelnutPOPL}. 
Hazel has both a textual syntax and a structure editor, which are briefly described in Sec.~\ref{sec:calculus-hazel}.\todo{sec} 
Marking is not integrated with structure editing, as it was in Hazelnut. 
%offering (1) a textual syntax, which allows the programmer to manually insert empty holes as necessary, (2) a text-like structure editor that offers partial syntax error recovery (inserting holes to maintain well-formedness as long as all matching delimiters have been placed) \cite{tylr}, and (3) a term-based structure editor with total syntax error recovery, i.e. a guarantee that every edit maintains syntactic well-formedness \cite{HazelnutPOPL}\todo{popl17}. 
However, as a secondary contribution, 
we briefly discuss in Sec.~\ref{sec:calculus-structured-editing} how (1)  total marking can resolve the problem of undefined editing behavior in the Hazelnut structure editor calculus, 
and (2) integrating marking with typed structure editing as specified by Hazelnut allows us to re-mark only where necessary based on the edit location and the type and binding structure.

Local type inference pairs well with type error localization and recovery because information flows systematically down the tree, leading to predictable and systematic localization decisions (as informally observed also by \cite{pierce}). Languages with constraint-based type inference gather constraints from all relevant locations in a program. This approach is powerful, but it is also notorious for complicating type error localization and recovery, because inconsistencies can arise through a confluence of constraints originating from any number of locations in the program. 
Complex techniques, e.g. based on machine learning or manual weighting, have been proposed to heuristically localize these errors to expressions. 
In Sec.~\ref{sec:thi}\todo{sec}, we describe how we blend local and constraint-based type inference via type holes in Hazel. In particular,  
we generate constraints for each type hole that appears in the program explicitly or that is  generated internally during the  bidirectional type error localization and recovery process from Sec.~\ref{sec:calculus}. 
We develop a unification algorithm that can recover from inconsistent constraints by maintaining possible type hole fillings. For holes that can be uniquely solved, we suggest the solution. 
If conflicting constraints arise, we localize the errors to the relevant type holes, rather 
than attempting to localize to expressions. 
The user can explore different partial type hole fillings. Making a choice allows the bidirectional system to shift the error to relevant expressions.
Until a choice is made, the system continues to recover gradually.
